XRONOS: an open data infrastructure for archaeological chronology

### 2024-11-29 paper.qmd
Add fig-data-model

```{r setup}
#| include: false
library("countrycode")
library("cowplot")
library("dplyr", warn.conflicts = FALSE)
library("giscoR")
library("ggplot2")
library("glue")
library("gt", warn.conflicts = FALSE)
library("khroma")
library("patchwork", warn.conflicts = FALSE)
library("readr")
library("sf")
library("tidyr")

@@

<!--
* [x] First draft
* [ ] Complete references:
  * [x] Introduction
  * [ ] State of the art
  * [ ] Concept
  * [ ] Implementation
  * [ ] Evaluation
  * [ ] Appendix table
* [ ] Figures and tables
  * [x] FIGURE: graph of number of datasets through time
  * [x] FIGURE: map of number of dates in datasets
  * [ ] FIGURE: distribution of radiocarbon dates in c14bazAAR / IntChron / p3k14c / XRONOS
  * [ ] FIGURE: data model
  * [ ] FIGURE: GUI screenshots (data browser, c14 show view, version history, data curation)
  * [ ] TABLE: list of 'issues'
  * [ ] APPENDIX: table of datasets
* [ ] Other TODOs
* [ ] Copyedit:
  * [ ] Introduction
  * [ ] State of the art
  * [ ] Concept
  * [ ] Implementation
  * [ ] Evaluation
  -->

@@

<!-- TODO: data model figure -->

@@

At the base of the XRONOS data model are sets of spatiotemporal coordinates or, as we call them, *chrons*.
In an archaeological context, we conceptualise a chron as an assertion linking human activity with a particular point in space and time.
Our data model currently encompasses three types of chron: radiocarbon dates, typological dates (e.g. 'Early Neolithic') and dendrochronological dates.
However we anticipate that the concept will accommodate other types of absolute and relative dating techniques, as the scope of the database expands.

>>>>

```{r setup}
#| include: false
library("countrycode")
library("cowplot")
library("dplyr", warn.conflicts = FALSE)
library("dm")
library("giscoR")
library("ggplot2")
library("glue")
library("gt", warn.conflicts = FALSE)
library("khroma")
library("magick")
library("patchwork", warn.conflicts = FALSE)
library("readr")
library("RPostgres")
library("sf")
library("tidyr")
library("webshot2")

@@

<!--
* [x] First draft
* [ ] Complete references:
  * [x] Introduction
  * [ ] State of the art
  * [ ] Concept
  * [ ] Implementation
  * [ ] Evaluation
  * [ ] Appendix table
* [ ] Figures and tables
  * [x] FIGURE: graph of number of datasets through time
  * [x] FIGURE: map of number of dates in datasets
  * [ ] FIGURE: distribution of radiocarbon dates in c14bazAAR / IntChron / p3k14c / XRONOS
  * [x] FIGURE: data model
  * [ ] FIGURE: GUI screenshots (data browser, c14 show view, version history, data curation)
  * [ ] TABLE: list of 'issues'
  * [ ] APPENDIX: table of datasets
* [ ] Other TODOs
* [ ] Copyedit:
  * [ ] Introduction
  * [ ] State of the art
  * [ ] Concept
  * [ ] Implementation
  * [ ] Evaluation
  -->

@@

```{r fig-data-model}
#| fig-cap: Simplified entity relationship diagram showing the XRONOS data model
# Requires a running XRONOS dev database
# See https://github.com/xronos-ch/xronos.rails for setup instructions
xronos_dev_db <- DBI::dbConnect(
  RPostgres::Postgres(),
  user= "xronos",
  password = "xronos",
  dbname = "xronos_dev",
  host = "localhost"
)

xronos_dm <- dm_from_con(xronos_dev_db, learn_keys = TRUE) # doesn't learn much

xronos_dm_svg <- xronos_dm |>
  dm_select_tbl(c("c14_labs", "c14s", "cals", "citations", "contexts",
                  "materials", "references", "samples", "site_names",
                  "site_types", "site_types_sites", "sites", "taxons", "typos", 
                  "users", "versions", "wikidata_links")) |>
  # Relations
  dm_add_fk("c14s", "c14_lab_id", "c14_labs") |>
  dm_add_fk("c14s", "sample_id", "samples") |>
  dm_add_fk("cals", c("c14_age", "c14_error"), "c14s", c("bp", "std")) |>
  dm_add_fk("citations", "reference_id", "references") |>
  dm_add_fk("citations", "citing_id", "sites") |>
  dm_add_fk("citations", "citing_id", "c14s") |>
  dm_add_fk("citations", "citing_id", "typos") |>
  dm_add_fk("contexts", "site_id", "sites") |>
  dm_add_fk("samples", "context_id", "contexts") |>
  dm_add_fk("samples", "material_id", "materials") |>
  dm_add_fk("samples", "taxon_id", "taxons") |>
  # dm_add_fk("site_names", "site_id", "sites") |> # autodetected
  dm_add_fk("site_types_sites", "site_id", "sites") |>
  dm_add_fk("site_types_sites", "site_type_id", "site_types") |>
  dm_add_fk("typos", "sample_id", "samples") |>
  dm_add_fk("wikidata_links", "wikidata_linkable_id", "sites") |>
  dm_add_fk("wikidata_links", "wikidata_linkable_id", "c14s") |>
  dm_add_fk("wikidata_links", "wikidata_linkable_id", "c14_labs") |>
  dm_add_fk("wikidata_links", "wikidata_linkable_id", "materials") |>
  dm_add_fk("wikidata_links", "wikidata_linkable_id", "taxons") |>
  dm_add_fk("versions", "whodunnit", "users", "id") |>
  dm_add_fk("versions", "item_id", "c14_labs") |>
  dm_add_fk("versions", "item_id", "c14s") |>
  dm_add_fk("versions", "item_id", "contexts") |>
  dm_add_fk("versions", "item_id", "materials") |>
  dm_add_fk("versions", "item_id", "references") |>
  dm_add_fk("versions", "item_id", "samples") |>
  dm_add_fk("versions", "item_id", "site_names") |>
  dm_add_fk("versions", "item_id", "site_types") |>
  dm_add_fk("versions", "item_id", "sites") |>
  dm_add_fk("versions", "item_id", "taxons") |>
  dm_add_fk("versions", "item_id", "typos") |>
  # TODO: Self-references (primarily persuadable models), or too messy?
  # dm_add_fk("contexts", "superseded_by", "contexts") |>
  dm_draw("BT", view_type = "title_only", column_types = TRUE) |>
  DiagrammeRsvg::export_svg()

# Dear god why
xronos_dm_svg_file <- tempfile("xronos_dm_", fileext = ".svg")
write(xronos_dm_svg, xronos_dm_svg_file)
ggdraw() + draw_image(xronos_dm_svg_file)
```

@@

At the base of the XRONOS data model (@fig-data-model) are sets of spatiotemporal coordinates or, as we call them, *chrons*.
In an archaeological context, we conceptualise a chron as an assertion linking human activity with a particular point in space and time.
Our data model currently encompasses three types of chron: radiocarbon dates, typological dates (e.g. 'Early Neolithic') and dendrochronological dates.
However we anticipate that the concept will accommodate other types of absolute and relative dating techniques, as the scope of the database expands.
